How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-08 doc/paper.tex
kept hacking away at abstract

  One of the central tenants of conservation paleobiology is that by studying the past we can better predict the future. It has been suggested that risk assessments for extant species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. Such suggestions prompt two key questions: (1) At a given timescale, to what extent are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) At a given point in time, how accurate do extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns  turn out to be? To address these questions we analyze the exceptionally well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, climate state, and the change in climate state. Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

>>>>

  One of the central tenants of conservation paleobiology is that risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. 
  We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
  To address these questions we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, recent change in geographic range, climate state, and the change in climate state. Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
  %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) 
  %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
